Here's the updated `supply-demand-equilibrium.qmd` with a new **Graph Paper Grids** section and ggplot2 code for background grids:

```qmd
---
title: "Linear Demand and Supply Models"
subtitle: "Based on Cowen & Tabarrok MRU Principles of Microeconomics, Ch. 3–4"
author: "Professor [Your Name]"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)

theme_set(
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(alpha = 0.3),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold")
  )
)

demand_color      <- "#2E86AB"
supply_color      <- "#A23B72"
equilibrium_color <- "#F18F01"
surplus_color     <- "#C73E1D"

grid_color        <- "#CCCCCC"
create_demand <- function(a, b) function(p) pmax(0, a - b * p)
create_supply <- function(c, d) function(p) pmax(0, c + d * p)
```

# Graph Paper Grids

This section overlays demand and supply curves onto a graph-paper style grid (light gray major and minor grid lines) to aid students in reading values directly from the plot.

```{r grid-plot, fig.height=5, fig.width=6}
# Sample curves for demonstration
 demand <- create_demand(120, 4)
 supply <- create_supply(-20, 6)
 p_vals <- seq(0, 35, 0.5)
 df_grid <- tibble(
   P = p_vals,
   Qd = demand(p_vals),
   Qs = supply(p_vals)
 )

 ggplot(df_grid, aes(x = Qd, y = P)) +
   # Graph-paper grid with major and minor lines
   geom_hline(yintercept = seq(0, 35, 5), color = grid_color, size = 0.3) +
   geom_vline(xintercept = seq(0, 140, 20), color = grid_color, size = 0.3) +
   geom_hline(yintercept = seq(0, 35, 1), color = grid_color, linetype = "dotted", size = 0.2) +
   geom_vline(xintercept = seq(0, 140, 5), color = grid_color, linetype = "dotted", size = 0.2) +
   # Demand and supply curves
   geom_line(aes(x = Qd, y = P), color = demand_color, size = 1.2) +
   geom_line(data = df_grid, aes(x = Qs, y = P), color = supply_color, size = 1.2) +
   # Equilibrium point
   geom_point(aes(x = 64, y = 14), color = equilibrium_color, size = 3) +
   annotate("text", x = 64, y = 14 + 2, label = "(64,14)", color = equilibrium_color, fontface = "bold") +
   labs(
     title = "Supply and Demand with Graph-Paper Grid",
     x = "Quantity (Q)",
     y = "Price (P)"
   ) +
   xlim(0, 140) + ylim(0, 35)
```

<!-- Append any additional content sections below -->

```
