---
title: "Deriving Demand for Rum and Swords"
---

## Introduction {.center}

🎯 **Learning Objective**

Derive individual demand curves from utility maximization theory

**Our Example:** Two goods - 🍺 **Rum** (x-axis) - ⚔️ **Swords** (y-axis)

## The Budget Constraint {.smaller}

### Basic Setup

Consumer income: $M$ to spend on two goods: - **Rum** ($R$) with price $p_R$ - **Swords** ($S$) with price $p_S$

### Budget Constraint Equation

$$p_R \cdot R + p_S \cdot S = M$$

### Slope-Intercept Form

$$S = \frac{M}{p_S} - \frac{p_R}{p_S} \cdot R$$

## Initial Scenario {.center}

### Our Assumptions

-   **Income:** $M = \$120$
-   **Price of Rum:** $p_R = \$4$ per bottle\
-   **Price of Swords:** $p_S = \$6$ per sword

### Initial Budget Constraint

$$4R + 6S = 120$$

### Slope-Intercept Form

$$S = 20 - \frac{2}{3}R$$

## Utility Function {.center}

### Cobb-Douglas Utility Function

$$U(R,S) = R^{0.6} \cdot S^{0.4}$$

**Why Cobb-Douglas?** - Mathematically tractable - Realistic diminishing marginal utility - Widely used in economics

## Utility Maximization {.smaller}

### The Key Condition

Marginal Rate of Substitution = Price Ratio

$$MRS_{R,S} = \frac{MU_R}{MU_S} = \frac{p_R}{p_S}$$

### For Cobb-Douglas Function

-   $MU_R = 0.6 \cdot R^{-0.4} \cdot S^{0.4}$
-   $MU_S = 0.4 \cdot R^{0.6} \cdot S^{-0.6}$

### Therefore:

$$MRS_{R,S} = 1.5 \cdot \frac{S}{R}$$

## Finding the Optimal Bundle {.center}

### Setting MRS = Price Ratio

$$1.5 \cdot \frac{S}{R} = \frac{4}{6} = \frac{2}{3}$$

### Solving for S in terms of R

$$S = \frac{4}{9}R$$

### Substituting into Budget Constraint

$$4R + 6 \cdot \frac{4}{9}R = 120$$

### **Result:** $(R^*, S^*) = (18, 8)$

## Price Changes: Rum = \$3 {.center}

### New Price Ratio

$$1.5 \cdot \frac{S}{R} = \frac{3}{6} = \frac{1}{2}$$

### New Relationship

$$S = \frac{1}{3}R$$

### New Budget Constraint

$$3R + 6S = 120$$

### **New Optimum:** $(R^*, S^*) = (24, 8)$

## Price Changes: Rum = \$6 {.center}

### New Price Ratio

$$1.5 \cdot \frac{S}{R} = \frac{6}{6} = 1$$

### New Relationship

$$S = \frac{2}{3}R$$

### **New Optimum:** $(R^*, S^*) = (12, 8)$

## Price Changes: Rum = \$2 {.center}

### Following the same process...

### **New Optimum:** $(R^*, S^*) = (36, 8)$

**Notice:** Swords consumption stays constant at 8!

## Demand Schedule {.center}

| Price of Rum | Quantity of Rum | Quantity of Swords | Total Utility |
|:-------------|:----------------|:-------------------|:--------------|
| \$6          | 12 bottles      | 8 swords           | 57.9          |
| \$4          | 18 bottles      | 8 swords           | 67.2          |
| \$3          | 24 bottles      | 8 swords           | 73.8          |
| \$2          | 36 bottles      | 8 swords           | 85.7          |

## Mathematical Demand Function {.center}

### General Cobb-Douglas Result

For $a = 0.6, b = 0.4, M = 120, p_S = 6$:

### **Rum Demand Function:**

$$R^*(p_R) = \frac{72}{p_R}$$

### **Swords Demand Function:**

$$S^* = 8$$

## Demand Curve Visualization {.center}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)

# Create data for demand curve
price_range <- seq(1, 8, 0.1)
quantity_demanded <- 72 / price_range

demand_curve_data <- data.frame(
  Price = price_range,
  Quantity = quantity_demanded
)

# Specific points from our analysis
points_data <- data.frame(
  Price = c(6, 4, 3, 2),
  Quantity = c(12, 18, 24, 36)
)

# Create the plot
ggplot() +
  geom_line(data = demand_curve_data, aes(x = Quantity, y = Price), 
            color = "#4ECDC4", linewidth = 2) +
  geom_point(data = points_data, aes(x = Quantity, y = Price), 
             color = "#FF6B6B", size = 6) +
  geom_text(data = points_data, aes(x = Quantity, y = Price, 
                                    label = paste("($", Price, ",", Quantity, ")")), 
            vjust = -1, hjust = 0.5, size = 4, color = "white") +
  labs(title = "Demand Curve for Rum",
       subtitle = "R* = 72/p_R",
       x = "Quantity of Rum (bottles)",
       y = "Price of Rum ($)") +
  theme_dark() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold", color = "white"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, color = "#4ECDC4"),
    axis.title = element_text(size = 14, color = "white"),
    axis.text = element_text(size = 12, color = "white"),
    panel.background = element_rect(fill = "transparent"),
    plot.background = element_rect(fill = "transparent")
  ) +
  scale_x_continuous(breaks = seq(0, 40, 5), limits = c(0, 40)) +
  scale_y_continuous(breaks = seq(0, 8, 1), limits = c(0, 8))
```

## Budget Lines Visualization {.center}

```{r}
#| echo: false
#| message: false
#| warning: false

# Create budget line data
rum_range <- seq(0, 40, 1)

# Budget lines for different prices
budget_long <- data.frame(
  Rum = rep(rum_range, 4),
  Swords = c(20 - rum_range, 20 - (2/3) * rum_range, 
             20 - (1/2) * rum_range, 20 - (1/3) * rum_range),
  Price = rep(c("$6", "$4", "$3", "$2"), each = length(rum_range))
)

# Filter valid points
budget_long <- budget_long[budget_long$Swords >= 0, ]

# Optimal points
optimal_points <- data.frame(
  Rum = c(12, 18, 24, 36),
  Swords = c(8, 8, 8, 8),
  Price = c("$6", "$4", "$3", "$2")
)

# Create the plot
ggplot() +
  geom_line(data = budget_long, aes(x = Rum, y = Swords, color = Price), 
            linewidth = 2) +
  geom_point(data = optimal_points, aes(x = Rum, y = Swords, color = Price), 
             size = 6) +
  labs(title = "Budget Constraints and Optimal Consumption",
       subtitle = "Income = $120, Price of Swords = $6",
       x = "Rum (bottles)",
       y = "Swords (quantity)",
       color = "Price of Rum") +
  theme_dark() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold", color = "white"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, color = "#4ECDC4"),
    axis.title = element_text(size = 14, color = "white"),
    axis.text = element_text(size = 12, color = "white"),
    legend.title = element_text(color = "white"),
    legend.text = element_text(color = "white"),
    panel.background = element_rect(fill = "transparent"),
    plot.background = element_rect(fill = "transparent")
  ) +
  scale_x_continuous(breaks = seq(0, 40, 5), limits = c(0, 40)) +
  scale_y_continuous(breaks = seq(0, 25, 5), limits = c(0, 25)) +
  scale_color_manual(values = c("$2" = "#1f77b4", "$3" = "#ff7f0e", 
                                "$4" = "#2ca02c", "$6" = "#d62728"))
```

## Key Economic Insights {.smaller}

### 📊 Elasticity Analysis

For demand function $R = \frac{72}{p_R}$:

$$\varepsilon_{R,p_R} = -1$$

**Unit Elastic Demand** at all points!

### 💡 Income and Substitution Effects

When price decreases from \$4 to \$3: - **Total Effect:** $\Delta R = 24 - 18 = 6$ bottles - Both effects reinforce the law of demand

## Applications {.center}

### 🏪 Business Applications

-   **Pricing strategies**
-   **Revenue optimization**\
-   **Market analysis**

### 🏛️ Policy Applications

-   **Tax policy effects**
-   **Subsidy programs**
-   **Market regulation**

## Key Takeaways {.center}

### ✅ What We Learned

1.  **Budget constraints** limit consumer choices
2.  **Utility maximization** determines optimal bundles\
3.  **Price changes** create predictable demand responses
4.  **Mathematical functions** can be derived analytically
5.  **Elasticity** provides insights into behavior

### 🎯 **The Framework**

Essential tools for economic analysis and policy evaluation

## Questions? {.center}

### 🤔 **Discussion Topics**

-   What happens with different utility function parameters?
-   How do income changes affect demand?
-   What about substitute and complement goods?

### 📚 **Next Steps**

-   Practice with problem sets
-   Explore interactive simulations
-   Apply to real-world examples
